% !TeX root = ./diss.tex

\documentclass[12pt,notitlepage]{report}

\usepackage{a4}
\usepackage{verbatim}
\usepackage[utf8]{inputenc}
\usepackage[british,UKenglish]{babel}
\usepackage{enumitem}
\usepackage{parskip}
\usepackage{biblatex}
\usepackage{float}
\usepackage{placeins}
\usepackage{csquotes}
\usepackage{listings}
\usepackage{dirtree}
\usepackage{float}
\usepackage{caption}
\usepackage{subcaption}
\usepackage[binary-units=true]{siunitx}
\usepackage[final]{pdfpages}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage{fancyhdr}
\usepackage[raggedright]{titlesec}
\usepackage[a4paper,left=2cm,right=2cm,top=2.5cm,bottom=2cm]{geometry}% margins
\usepackage{hyperref}
\usepackage{cleveref}

\addbibresource{refs.bib}
\defbibheading{secbib}[\bibname]{%
  \chapter*{References}%
  \markboth{#1}{#1}}

\titleformat{\chapter}[hang]    % Chapter number and name on one line
{\normalfont\huge\bfseries}{\chaptertitlename\ \thechapter:}{1em}{}
\titlespacing*{\chapter}{0pt}{0pt}{6pt}

\setcounter{secnumdepth}{4}     % subsubsections (H3) enumerated
\setcounter{tocdepth}{4}        % subsubsections (H3) in table of contents

% Fancy section referencing
\crefformat{section}{\S#2#1#3}
\crefformat{subsection}{\S#2#1#3}
\crefformat{subsubsection}{\S#2#1#3}
\crefrangeformat{section}{\S\S#3#1#4 to~#5#2#6}
\crefmultiformat{section}{\S\S#2#1#3}{ and~#2#1#3}{, #2#1#3}{ and~#2#1#3}

\definecolor{darkorange}{rgb}{0.82, 0.41, 0.12}
\definecolor{darkblue}{rgb}{0.0, 0.28, 0.67}

\DeclareSIUnit{\bits}{bits}
\DeclareSIUnit{\cycle}{/ cycle}
\DeclareSIUnit{\cycles}{cycles}
\sisetup{group-separator = {,}}

\newcommand{\candidate}{2398E}

\lstset{
  basicstyle=\ttfamily
}

\raggedbottom                           % try to avoid widows and orphans
\sloppy
\clubpenalty1000%
\widowpenalty1000%

\renewcommand{\baselinestretch}{1.1}    % adjust line spacing to make
                                        % more readable

\begin{document}

%\bibliographystyle{plain}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title


\pagestyle{empty}

\hfill{\LARGE \bf Benjamin Marshall}

\vspace*{60mm}
\begin{center}
\Huge
{\bf Parallelising Sequence Alignment} \\
\vspace*{5mm}
\LARGE
Computer Science Tripos -- Part II \\
\vspace*{5mm}
Sidney Sussex College \\
\vspace*{5mm}
2020  % today's date
\end{center}

\newpage

\pagestyle{plain}
\pagenumbering{arabic}
\setcounter{page}{2}

\section*{Declaration}

I, Benjamin Marshall of Sidney Sussex College, being a candidate for Part II of
the Computer Science Tripos, hereby declare that this dissertation and the work
described in it are my own work, unaided except as may be specified below, and
that the dissertation does not contain material that has already been used to
any substantial extent for a comparable purpose.

I, Benjamin Marshall of Sidney Sussex College, am content for my dissertation to
be made available to the students and staff of the University.

\bigskip
\leftline{Signed}

\medskip
\leftline{Date \today}

\section*{Acknowledgements}

I would like to thank my supervisor Peter Rugg for his guidance and feedback throughout the project, and my Director of Studies Matthew Ireland for his feedback on this dissertation.

\newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Proforma, table of contents and list of figures

\chapter*{Proforma}

{\large
\begin{tabular}{ll}
Candidate Number:   & \bf \candidate                              \\
Project Title:      & \bf Parallelising Sequence Alignment        \\
Examination:        & \bf Computer Science Tripos -- Part II      \\
Examination Year:   & \bf 2020                                    \\
Word Count:         & \bf 11,945\footnotemark[1]                    \\
Line Count:         & \bf 5,172\footnotemark[2]                    \\
Project Originator: & Mr P.~Rugg                    \\
Supervisor:         & Mr P.~Rugg                    \\
\end{tabular}
}
\footnotetext[1]{
  This word count was computed by
  \lstinline{detex -l chapters/*.tex | tr -cd '0-9A-Za-z \\n' | wc -w}
}
\stepcounter{footnote}
\footnotetext[2]{
  This line count was computed using \lstinline{cloc --exclude-dir=quartus}; it excludes comments and blank lines.
}
\stepcounter{footnote}


\section*{Original Aims of the Project}

To implement the Smith-Waterman algorithm for sequence alignment in single-threaded C, multi-threaded C, and CUDA, and to thoroughly evaluate their performance.
One extension was to modify the C implementation to use Streaming SIMD Extensions to parallelise that program further.
The other extension was to implement the Smith-Waterman algorithm in SystemVerilog and instantiate it on an FPGA.
Each implementation was to utilise a variety of alignment-scoring approaches.

\section*{Work Completed}

The Smith-Waterman algorithm was implemented in single-threaded C, multi-threaded C, and CUDA (satisfying the core success criteria), and also in SystemVerilog.
The SystemVerilog implementation is limited to smaller sequences than the C and CUDA implementations.
Their performance was compared against each other, and across different alignment-scoring metrics as well.

\section*{Special Difficulties}

None.

\newpage

\tableofcontents

% \listoffigures

\newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% now for the chapters

\pagestyle{fancy}
\renewcommand{\sectionmark}[1]{\markright{\textsl{\MakeUppercase{\thesection.\ #1}}}}
\lhead{\rightmark}
\rhead{}
\cfoot{\thepage}
\renewcommand{\headrulewidth}{0pt}

\include{chapters/intro}

\include{chapters/prep}

\include{chapters/impl}

\include{chapters/eval}

\include{chapters/concl}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% the bibliography

\addcontentsline{toc}{chapter}{Bibliography}
\printbibliography[heading=secbib]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% the appendices
\appendix

% \chapter{Repository overview}
% \label{sec:Repo_overview_app}

% \input{repo_overview}

\chapter{Project Proposal}
\label{sec:Proposal}

\input{propbody}

\end{document}
